---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r, echo=TRUE, warning=FALSE, message=FALSE}
#******************************************************************
#Step 0. Downloading dependencies
#******************************************************************
library(tidyverse)
library(readr)
library(BiocManager)
library(impute)

# 1. Download data in to data folder (git will ignore this folder)
if(!file.exists("./data")){dir.create("./data")}

if(!exists("activity")) {
  fileUrl<- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  
  zipF <- "./data/activity_monitoring_data.zip"
  
  if(!file.exists(zipF)) {
    download.file(fileUrl, destfile = "./data/activity_monitoring_data.zip", method = "curl")
    outDir <- "./data"
    activity <- unzip(zipF, exdir = outDir)
  }
  
  # 2. Read data into global environment
  activity <- read.table("data/activity.csv", header = TRUE, sep = ",", dec = ".")
}

# 3. Format the date variable with as.Date
activity$date <- as.Date(activity$date, "%Y-%m-%d")
activity$weekday <- weekdays(activity$date)

summary(activity)

```


## What is mean total number of steps taken per day?

```{r histogram}
hist_data <- aggregate(steps ~ date, data = activity, sum, na.rm = TRUE)

hist <- ggplot(hist_data, aes(x=steps)) + 
  geom_histogram(fill="#69b3a2", color="#e9ecef", binwidth = 2500) +
    ggtitle("Distribution of total number of steps taken per day") +
  ylab("Frequency") + 
  xlab("Number of steps")

hist

```

- The average number of steps taken per day: 
```{r, include=TRUE}
mean <- round(mean(hist_data$steps), 2)
mean
```

- The median number of steps taken per day: 

```{r, include=TRUE}
median <- median(hist_data$steps)
median
```

## What is the average daily activity pattern?

- Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  

```{r timeseries}
time_data <- aggregate(activity$steps, by=list(activity$interval), FUN=mean, na.rm=TRUE)
colnames(time_data) <- c("interval", "steps")

plot(time_data$interval, time_data$steps, type = "l", 
     main = "Average number of steps taken by 5-minute interval", 
              ylab = "Time (5-minute increments)", 
              xlab = "Average number of steps", 
              pch = 1)


```

- Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}

time_data[which.max(time_data$steps), ]$interval

```


## Imputing missing values  

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
summary <- summary(activity)
nas <- summary[7, 1]
nas <- str_split(nas, ":", simplify = TRUE)
nas <- as.numeric(nas[,2])
```
- The only variable with missing values in the original dataset is 'steps'. Within 'steps' there are `r nas` missing values. 

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  
  - Strategy: If 'steps' is missing, fill that value with the mean for that 5-minute interval. 
  
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activity_impute <- activity

for (n in 1:nrow(activity_impute)) {
  if(is.na(activity_impute$steps[n])){
    time_data_row_n <- (n%%288)
    if (time_data_row_n == 0) {
      time_data_row_n <- 288
    }
    activity_impute$steps[n] <- time_data$steps[time_data_row_n]
  } 
}
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
hist_data_impute <- aggregate(steps ~ date, data = activity_impute, sum, na.rm = TRUE)

hist_impute <- ggplot(hist_data_impute, aes(x=steps)) + 
  geom_histogram(fill="#69b3a2", color="#e9ecef", binwidth = 2500) +
    ggtitle("Distribution of total number of steps taken per day (imputed missing values)") +
  ylab("Frequency") + 
  xlab("Number of steps")

hist_impute

```

- The average number of steps taken per day with imputed data: 

```{r, echo=TRUE}
mean_imp <- round(mean(hist_data_impute$steps), 2)
mean_imp
```

- The median number of steps taken per day with imputed data: 

```{r}
median_imp <- median(hist_data_impute$steps)
median_imp
```
The average number of steps taken per day is `r mean_imp` and the median is `r median_imp`. 

## Are there differences in activity patterns between weekdays and weekends?
